---
title: "MLB run scoring trends: data wrangle"
author: "Martin Monkman"
date: "January 26, 2020"
output: html_document
---



This note provides a description of the steps I took to update my run scoring trends data visualization application, built using RStudio's shiny package.

You can find the app here:
[https://monkmanmh.shinyapps.io/MLBrunscoring_shiny/](https://monkmanmh.shinyapps.io/MLBrunscoring_shiny/)

The github repo for this app is [https://github.com/MonkmanMH/Bayesball/tree/master/MLBrunscoring_shiny](https://github.com/MonkmanMH/Bayesball/tree/master/MLBrunscoring_shiny)

This update gave me the opportunity to make some more cosmetic tweaks to the front end, and some subtle changes to the code under the hood.


My previous update switched from using the [{reshape}](https://cran.r-project.org/web/packages/reshape/index.html) package, and moved to [{tidyr}](https://tidyr.tidyverse.org/), part of [the tidyverse](https://www.tidyverse.org/). 



## 2. wrangling: the calculation of team and league run scoring averages

The goal is to create data tables that minimize the amount of processing the app has to do. What follows is a detailed explanation of how I approached this. 

It's always valuable to have your end-state in mind when working through a multi-step data wrangle like this. My goal is the values shown on the "team plot" tab of the app&mdash;an index value (i.e. a percentage) of a team's average runs scored (and runs allowed) compared to the league run scoring rate, for a single season.


### a. Load packages and read the data


First, load the necessary packages, the first four of which are part of the tidyverse.

```{r setup, warning=FALSE, message=FALSE}

# tidyverse packages
library(dplyr)
library(purrr)
library(readr)
library(tidyr)

library(feather)

```

Then, read in the data.

- It comes from the "Lahman" baseball database, so-named after the originator and host, Sean Lahman. [The website www.seanlahman.com](http://www.seanlahman.com/baseball-archive/statistics/) has the current version of the database, as well as an archive of previous versions.

- The data table "Teams" is sourced from [the _baseballdatabank_ on github] (https://github.com/chadwickbureau/baseballdatabank); I've also saved the file in the github repo for this app.

- There is an [R package with the Lahman database](https://github.com/cdalzell/Lahman), but beware: the data in the CRAN version does not have the most recent season.


```{r}

Teams <- read_csv("Teams.csv",
                  col_types = cols(
                    divID = col_character(),
                    DivWin = col_character(),
                    SF = col_character(),
                    WCWin = col_character()
                  ))

Teams

```


The table above has far more variables than what we need, and some that we'll have to calculate. 



### b. Create league summary tables

A short set of instructions that starts with the "Teams" table in the Lahman database
and summarizes it for [MLB run scoring trends](https://monkmanmh.shinyapps.io/MLBrunscoring_shiny/) Shiny app

Thus rather than having the app do the work of

1. remove unnecessary records (rows) and fields (columns) and

2. run the calculations for the runs-per-game, runs-allowed-per-game, and indexed versions of those,

the calculations are conducted here. This will vastly improve the performance of the app.


#### i. create nested table


I started with the "Many Models"" chapter of Wickham and Grolemund, [R for Data Science](https://r4ds.had.co.nz/many-models.html). (And thanks to Dr. Charlotte Wickham, whose training course was invaliable in helping me wrap my head around this.)

At this point, the code 

- filters out the years prior to 1901 and [the misbegotten Federal League](https://en.wikipedia.org/wiki/Federal_League).

- and then creates a nested data table, starting with the `group_by()` year and league (`lgID`). This table has three columns: the year, the league, and then the third column is a tibble, itself containing all the variables for each team.


```{r}

# select a sub-set of teams from 1901 [the establishment of the American League] forward to most recent year
Teams_lgyr <- Teams %>%
  filter(yearID > 1900, lgID != "FL") %>%
  group_by(yearID, lgID) %>%
  nest()

Teams_lgyr

```


Here's a quick peek inside the first entry of the "data" column...the American League, 1901.

```{r}

Teams_lgyr$data[[1]]


```



#### ii - functional programming

This step creates a league run scoring function, and then applies that to each tibble in the `data` column, using the `purrr::map()` function.

**Note:**

- In the gapminder example in _R for Data Science_, the variables were called using their names. In this case, for a reason I have not yet determined, we have to specify the data object they are coming from; e.g. for the runs variable `R`, we have to use `df$R` (not just `R`). 


First, a simple test, calculating runs scored, and checking to see if we got the right answer, b comparing that to the value calculated using `dplyr`:

```{r}

# base R format
leagueRuns_fun <- function(df) {
  sum(data = df$R)
}
league_year_runs <- map(Teams_lgyr$data, leagueRuns_fun)
league_year_runs[[1]]
#check the answer by old school `dplyr` method
Teams %>%
  filter(yearID == 1901,
         lgID == "AL") %>%
  summarise(leagueruns = sum(R))

```

Now we move on to the calculation of league averages.

For the first approach, the sum calculation is part of the function.

- There are two functions, one for Runs and the other for Runs Allowed. This is because I have not yet figured out how to specify two different variables (i.e. the name of the data object and the variable to be used in the function) in the map_() function and successfully have them carried into my calculation functions


- Also note that in order to be consistent with other sources, the number of games played is calculated using the sum of wins (`W`) and losses (`L`), rather than the number of games reported in the `G` variable. 



```{r}

# functions
leagueRPG_fun <- function(df) {
  sum(data = df$R) / (sum(data = df$W) + sum(data = df$L))
  }
leagueRAPG_fun <- function(df) {
  sum(data = df$RA) / (sum(data = df$W) + sum(data = df$L))
  }
# simple `map` version
league_year_RPG <- map(Teams_lgyr$data, leagueRPG_fun)
# embed as new columns in nested data object
Teams_lgyr <- Teams_lgyr %>%
  mutate(lgRPG = map_dbl(Teams_lgyr$data, leagueRPG_fun),
         lgRAPG = map_dbl(Teams_lgyr$data, leagueRAPG_fun))
Teams_lgyr

```


In the second approach:

- the league and year total runs, runs allowed, and games are first calculated using separate functions

- RPG and RAPG for each league and year combination are then calculated outside the nested tibbles



```{r}

# more functions - individual league by year totals
leagueR_fun <- function(df) {
  sum(data = df$R)
}

leagueRA_fun <- function(df) {
  sum(data = df$RA)
}

leagueG_fun <- function(df) {
  (sum(data = df$W) + sum(data = df$L))
}


Teams_lgyr <- Teams_lgyr %>%
  mutate(lgR = map_dbl(Teams_lgyr$data, leagueR_fun),
         lgRA = map_dbl(Teams_lgyr$data, leagueRA_fun),
         lgG = map_dbl(Teams_lgyr$data, leagueG_fun))


Teams_lgyr <- Teams_lgyr %>%
  mutate(lgRPG = (lgR / lgG),
         lgRAPG = (lgRA / lgG))

Teams_lgyr


```



#### iii. save LG_RPG files

And then write csv and feather versions. As noted above, the shiny app now uses the `feather` format.

Notes:

- rounding of variables _after_ all the calculations, simply to make the tables as viewed more legible.

- renaming of variables to correspond with shiny app names.


```{r}

LG_RPG <- Teams_lgyr %>%
  mutate(lgRPG = round(lgRPG, 2),
         lgRAPG = round(lgRAPG, 2)) %>%
  select(yearID, lgID, R = lgR, RA = lgRA, G = lgG, 
         leagueRPG = lgRPG, leagueRAPG = lgRAPG)

LG_RPG

write_csv(LG_RPG, "LG_RPG.csv")
write_feather(LG_RPG, "LG_RPG.feather")

```


### c. Repeat for MLB total

This only differs from the league summaries in the level of nesting; instead of grouping by year and league, it's only year  (`yearID`).

```{r}

Teams_lgyr <- Teams_lgyr %>%
  unnest() %>%
  group_by(yearID) %>%
  nest()

Teams_lgyr

Teams_lgyr <- Teams_lgyr %>%
  mutate(mlbR = map_dbl(Teams_lgyr$data, leagueR_fun),
         mlbRA = map_dbl(Teams_lgyr$data, leagueRA_fun),
         mlbG = map_dbl(Teams_lgyr$data, leagueG_fun),
         mlbRPG = (mlbR / mlbG),
         mlbRAPG = (mlbRA / mlbG))

Teams_lgyr

```


And again, we save the files for use in the shiny app.

```{r}

MLB_RPG <- Teams_lgyr %>%
  mutate(mlbRPG = round(mlbRPG, 2),
         mlbRAPG = round(mlbRAPG, 2)) %>%
  select(yearID, R = mlbR, RA = mlbRA, G = mlbG, 
         leagueRPG = mlbRPG, leagueRAPG = mlbRAPG)

write_csv(MLB_RPG, "MLB_RPG.csv")
write_feather(MLB_RPG, "MLB_RPG.feather")

```





### d. Individual team values

Calculate index of team run scoring against league average 

Note that we start with `unnest()` and create a new object, `Teams_append` ... a tibble with all of the variables exposed.


```{r}

Teams_append <- Teams_lgyr %>%
  unnest() %>%
  mutate(teamRPG=(R / (W + L)), 
         teamRAPG=(RA / (W + L)), 
         WLpct=(W / (W + L))) %>%
  # runs scored index where 100=the league average for that season
  mutate(R_index = (teamRPG / lgRPG) * 100) %>%
  mutate(R_index.sd = sd(R_index)) %>%
  mutate(R_z = (R_index - 100) / R_index.sd) %>%
  # runs allowed
  mutate(RA_index = (teamRAPG / lgRAPG) * 100) %>%
  mutate(RA_index.sd = sd(RA_index)) %>%
  mutate(RA_z = (RA_index - 100) / RA_index.sd) 


Teams_append

```



In this the final step, we first create a new data object `Teams_merge`.

Notes:

- rounding of a variety of the calculated variables, to address readability concerns.

- selection and renaming of variables to correspond with shiny app names.

- then write csv and feather versions. 

```{r}

Teams_merge <- Teams_append %>%
  mutate(lgRPG = round(lgRPG, 2),
         lgRAPG = round(lgRAPG, 2),
         WLpct = round(WLpct, 3),
         teamRPG = round(teamRPG, 2),
         teamRAPG = round(teamRAPG, 2),
         R_index = round(R_index, 1),
         RA_index = round(RA_index, 1)
         ) %>%
  select(yearID, lgID, franchID, teamID, name,
         W, L, WLpct, R.x = R, RA.x = RA, 
         teamRPG, leagueRPG = lgRPG, R_index,
         teamRAPG, leagueRAPG = lgRAPG, RA_index)

Teams_merge

write_csv(Teams_merge, "Teams_merge.csv")
write_feather(Teams_merge, "Teams_merge.feather")

```


And the feather files can now be incorproated into the shiny app.


-30-

