---
title: "Run scoring trends - references"
author: "Martin Monkman"
date: "Wednesday, November 12, 2014"
output: html_document
---

## Major League Baseball run scoring trends
### Version 2 (2015-01-04)
### Documentation

This app was written by Martin Monkman using the R package [Shiny](http://shiny.rstudio.com/).

All of the files associated with this app can be found on [github.com](github.com), 
at [MonkmanMH/MLBrunscoring_shiny](https://github.com/MonkmanMH/MLBrunscoring_shiny).


### References

**shiny**:  the RStudio site provides a wealth of materials for Shiny developers.

+ [tutorial](http://shiny.rstudio.com/tutorial/) -- a great place if you're new to Shiny, but a useful reference if you are looking for code examples too.
+ [The Shiny cheat sheet](http://shiny.rstudio.com/articles/cheatsheet.html) -- once you are rolling, this is a handy reference to have on your desk.
+ [widget gallery](http://shiny.rstudio.com/gallery/widget-gallery.html) -- with the wide range of UI widgets within Shiny, each with a variety of options, this gallery is an amazing resource.
+ [function reference](http://shiny.rstudio.com/reference/shiny/latest/) -- just like it says. Essential.  

**shiny - other sources**

+ [shiny tutorial using Gapminder data](https://stat545-ubc.github.io/shiny03_activity.html) -- a tutorial from the UBC course Stat 545 by Julia Gustavsen.


**dplyr**

+ [on joins](http://stackoverflow.com/questions/1299871/how-to-join-data-frames-in-r-inner-outer-left-right)


### The Code

#### server.R

The first few lines call the relevant packages required to run the app.  While most are functional, the Lahman package contains the data that will be used, specifically the "Teams" data set.  The Lahman package bundles all of the tables provided in the [Lahman database](http://www.seanlahman.com/baseball-archive/statistics/).

```{r}

# package load 
library(shiny)
library(dplyr)
library(ggplot2)
library(Lahman)
#
# CREATE LEAGUE SUMMARY TABLES
# ============================
#
# load the Lahman data table "Teams"
data(Teams)
#
# select a sub-set of teams from 1901 [the establishment of the American League] forward to most recent year
#
LG_RPG <- Teams %>%
  filter(yearID > 1900, lgID != "FL") %>%
  group_by(yearID, lgID) %>%
  summarise(R=sum(R), RA=sum(RA), G=sum(G)) %>%
  mutate(RPG=R/G, RAPG=RA/G)
# and a version with just the MLB totals
MLB_RPG <- Teams %>%
  filter(yearID > 1900, lgID != "FL") %>%
  group_by(yearID) %>%
  summarise(R=sum(R), RA=sum(RA), G=sum(G)) %>%
  mutate(RPG=R/G, RAPG=RA/G)
#
firstyear <- MLB_RPG$yearID[1]
mostrecentyear <- tail(MLB_RPG$yearID, 1)
#
#
# ####################################################
# 
# Define server logic
shinyServer(function(input, output) {
  
  # trendline select (TRUE / FALSE)
  output$trendlineselectvalue <- renderPrint({ input$trendlineselect })
  
  # trendline sensitivity (slider)
  output$trendline_sen <- renderPrint({ input$trendline_sen_sel })
  
  # trendline confidence interval (slider)
  output$trendline_conf <- renderPrint({ input$trendline_conf_sel })
  
# define yearrange for UI 
  output$lg_yearrange <- renderUI({
    sliderInput("lg_yearrange_input", label = h4("Select year range to plot"), 
                min = firstyear, max = mostrecentyear, value = c(firstyear, mostrecentyear), 
                step = 1, round = TRUE, format = "####",
                animate = TRUE)
  })
  #
# +++++ PLOTS: RUNS SCORED PER GAME BY LEAGUE

output$plot_MLBtrend <- renderPlot({
    # plot the data
    MLBRPG <- ggplot(MLB_RPG, aes(x=yearID, y=RPG)) +
             geom_point() +
             xlim(input$lg_yearrange_input[1], input$lg_yearrange_input[2]) +
             ylim(3, 6) +
             ggtitle(paste("Major League Baseball: runs per team per game", 
                           input$lg_yearrange_input[1], "-", input$lg_yearrange_input[2])) +
             xlab("year") + ylab("runs per game")  

    # plot each league separately?
    if (input$leaguesplitselect == TRUE) {
      MLBRPG <- ggplot(LG_RPG, aes(x=yearID, y=RPG)) +
        geom_point() +
        xlim(input$lg_yearrange_input[1], input$lg_yearrange_input[2]) +
        ylim(3, 6) +
        ggtitle(paste("Major League Baseball: runs per team per game", 
                      input$lg_yearrange_input[1], "-", input$lg_yearrange_input[2])) +
        xlab("year") + ylab("runs per game") +
        facet_grid(lgID ~ .)
    }
    # add trend line to plot?
      if (input$trendlineselect == TRUE) {
          MLBRPG <- MLBRPG + 
            stat_smooth(method=loess, 
                        span=input$trendline_sen_sel,
                        level=as.numeric(input$trendline_conf_sel))
         }
    # final plot
    MLBRPG

  })
  # ----------- end MLBtrend plot

#
})
# ----------- end shinyServer function 

```

#### ui.R

```{r}

# Run scoring in Shiny -- ui
#
# written by Martin Monkman
# last update: 2014-11-09
#
# resources used & plagarized
#  - RStudio's widget gallery http://shiny.rstudio.com/gallery/widget-gallery.html
#
library(shiny)

shinyUI(navbarPage("MLB run scoring trends",
                   tabPanel("league",
                            #
                            titlePanel("Per-game run scoring by league"),
                            #
                            # Sidebar with a dropdown list of ministry names
                            sidebarLayout(
                              
                              sidebarPanel(

                                # slider for year range
                                uiOutput("lg_yearrange"),
                                hr(),
                                # checkbox for league split into facet
                                checkboxInput("leaguesplitselect", label = h4("Plot each league separately"), value = FALSE),
                                hr(),
                                hr(),
                                # checkbox for trend line
                                checkboxInput("trendlineselect", label = h4("Add a trend line"), value = FALSE),
                                # slider bar for trend line sensitivity 
                                sliderInput("trendline_sen_sel", 
                                            label = (h5("Select trend line sensitivity")), 
                                            min = 0.05, max = 1, value = .50, step = .05),
                                hr(),
                                # radio buttons for trend line confidence interval 
                                radioButtons("trendline_conf_sel", 
                                             label = h5("Select trend line confidence level"),
                                             choices = list("0.10" = .1, "0.50" = 0.5,
                                                            "0.90" = 0.9, "0.95" = 0.95,
                                                            "0.99" = 0.99, "0.999" = 0.999, 
                                                            "0.9999" = 0.9999), 
                                             selected = 0.95),
                                hr(),
                                column(4, verbatimTextOutput("trendline_conf"))
                              ),
                              # ---- end sidebarPanel
                              # 
                              mainPanel(
                                plotOutput("plot_MLBtrend")
                              )
                              # ---- end mainPanel
                            )
                            # ---- end sidebarLayout                              
                   ),
                   # ---------- end tabPanel "league"
                   #
                   navbarMenu("More",
                              tabPanel("reference",
                                       mainPanel(
                                         includeMarkdown("runscoring_references.Rmd")
                                       )),
                              tabPanel("documentation",
                              mainPanel(
                                   includeMarkdown("runscoring_documentation.Rmd")
                            )
                   ))
                   
                   # ----- end tabPanel "reference"
))
# ---------- end navbarPage
#  
#  


```




*-30-*
