---
title: "MLB run scoring--summary tables"
author: "Martin Monkman"
date: "Sunday, March 01, 2015"
output: html_document
---

```{r, echo=FALSE}
# this chunk defines the variable `stepcounter` which is used to auto calculate the numbering for each heading
stepcounter <- 0
```


A short set of instructions that starts with the "Teams" table in the Lahman database and summarizes it for the Shiny app. 

Thus rather than having the app load the entire Lahman database package (`Lahman`) and then

1. remove unnecessary records (rows) and fields (columns) and
2. run the calculations for 
  + runs-per-game (RPG)
  + runs-allowed-per-game (RAPG), and
  + indexed versions of RPG and RAPG, 
  
the preliminary calculations are conducted here. By running the routine steps in advance, the performance of the app is improved.


### STEP `r (stepcounter <- stepcounter + 1)`: Package load


First, load the `dplyr` packages:

```{r, message=FALSE}

library(dplyr)

```

### STEP `r (stepcounter <- stepcounter + 1)`:



The next step is to create the league summary tables.  The source file is the "Teams" table from the Lahman database, in this instance from the .csv version.

(The Lahman database can be sourced at [http://www.seanlahman.com/baseball-archive/statistics/](http://www.seanlahman.com/baseball-archive/statistics/).)

Using the `dplyr` package, the code creates a new data frame `LG_RPG` by

* uses `filter` to select a sub-set of teams from 1901 (the establishment of the American League) forward to most recent year and omits the Federal League, 

* calculates the number of runs scored and games played (first `group_by` to define the league and year as the summary characteristics, and then `summarise` to calculate the results ), and

* finally then calculates the runs per game using `mutate`. 

The second block of code then repeats the process to create `MLB_RPG`, simply by removing leage (`lgID`) as one of the `group_by` variables.

```{r}

Teams <- read.csv("Teams.csv", header=TRUE)

LG_RPG <- Teams %>%
  filter(yearID > 1900, lgID != "FL") %>%
  group_by(yearID, lgID) %>%
  summarise(R=sum(R), RA=sum(RA), G=sum(G)) %>%
  mutate(leagueRPG=R/G, leagueRAPG=RA/G)
#
# and a version with just the MLB totals
MLB_RPG <- Teams %>%
  filter(yearID > 1900, lgID != "FL") %>%
  group_by(yearID) %>%
  summarise(R=sum(R), RA=sum(RA), G=sum(G)) %>%
  mutate(leagueRPG=R/G, leagueRAPG=RA/G)


```

### STEP `r (stepcounter <- stepcounter + 1)`:


Create a new version of table showing the results for each individual team, including adding the league  average scores calculated above.

```{r}

Teams.merge <-  Teams %>%
  mutate(teamRPG=(R/G), teamRAPG=(RA/G), WLpct=(W/G)) 
Teams.merge <- 
  merge(Teams.merge, LG_RPG, by = c("yearID", "lgID"))
#
# create new values to compare the individual team's runs/game compares to the league average that season
Teams.merge <- Teams.merge %>%
  # runs scored index where 100=the league average for that season
  mutate(R_index = (teamRPG/leagueRPG)*100) %>%
  mutate(R_index.sd = sd(R_index)) %>%
  mutate(R_z = (R_index - 100)/R_index.sd) %>%
  # runs allowed
  mutate(RA_index = (teamRAPG/leagueRAPG)*100) %>%
  mutate(RA_index.sd = sd(RA_index)) %>%
  mutate(RA_z = (RA_index - 100)/RA_index.sd)

```

### STEP `r (stepcounter <- stepcounter + 1)`:

A necessary step for which I have yet to find an easy way is to round the resulting calculated variables, so that the `Shiny` app's tables will only display two decimal places.

```{r}

LG_RPG$leagueRPG <- round(LG_RPG$leagueRPG, 2)  
LG_RPG$leagueRAPG <- round(LG_RPG$leagueRAPG, 2)  
#
MLB_RPG$leagueRPG <- round(MLB_RPG$leagueRPG, 2)  
MLB_RPG$leagueRAPG <- round(MLB_RPG$leagueRAPG, 2)  
#
Teams.merge$R_index <- round(Teams.merge$R_index, 2)  
Teams.merge$R_index.sd <- round(Teams.merge$R_index.sd, 2)  
Teams.merge$R_z <- round(Teams.merge$R_z, 2)  
Teams.merge$RA_index <- round(Teams.merge$RA_index, 2)  
Teams.merge$RA_index.sd <- round(Teams.merge$RA_index.sd, 2)  
Teams.merge$RA_z <- round(Teams.merge$RA_z, 2)  
Teams.merge$teamRPG <- round(Teams.merge$teamRPG, 2)  
Teams.merge$teamRAPG <- round(Teams.merge$teamRAPG, 2)  
Teams.merge$leagueRPG <- round(Teams.merge$leagueRPG, 2)  
Teams.merge$leagueRAPG <- round(Teams.merge$leagueRAPG, 2)  
Teams.merge$WLpct <- round(Teams.merge$WLpct, 3)  

#
tail(LG_RPG)
tail(MLB_RPG)
tail(Teams.merge, 2)

```

### STEP `r (stepcounter <- stepcounter + 1)`:

And finally, write the three tables as .csv files for use by the app.

```{r}

write.csv(LG_RPG, file="LG_RPG.csv", row.names=FALSE)
write.csv(MLB_RPG, file="MLB_RPG.csv", row.names=FALSE)
write.csv(Teams.merge, file="Teams_merge.csv", row.names=FALSE)

```



*-30-*